A free/open-source Kazakh-Tatar machine translation system

### 2013-04-17 paper/paper.tex
some examples from the bidix; add the number of stems

\subsection{Bilingual lexicon}
%FIXME: Ilnar, we need some good examples here
\begin{figure*}[htbp]
\begin{texttt}
    <e><p><l>\textbf{юнәлеш}<s n="n"/></l><r>\textbf{йүнәлеш}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{танау}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{морон}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{алы{\qipa ҫ}лы{\qipa ҡ}}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{йыра{\qipa ҡ}лы{\qipa ҡ}}<s n="n"/></r></p></e>
\end{texttt}
\caption{Example entries from the bilingual transfer lexicon. Tatar is on the left, and Bashkir on the right}
\label{fig:bidix}
%\vspace{-1em}
\end{figure*}

The bilingual lexicon currently contains 2,834 stem to stem correspondences and was build by hand by a bilingual 
speaker of Tatar and Bashkir, translating a frequency list of the Russian National Corpus\footnote{\url{http://ruscorpora.ru/en/}} into both languages in a spreadsheet. This spreadsheet was then converted into the Apertium
XML dictionary format.

>>>>

\subsection{Bilingual lexicon}
%FIXME: Ilnar, we need some good examples here
% ақпараттық құрал - мәгълүмат чарасы
% құрал - корал in other contexts

%<e>       <p><l>жай<s n="n"/></l>                 <r>җәя<s n="n"/></r></p></e> <!--"arch"-->
%<e>       <p><l>жай<s n="n"/></l>                 <r>урын<s n="n"/></r></p></e> <!--"place"-->
%<e>       <p><l>жай<s n="n"/></l>                 <r>хәл<s n="n"/></r></p></e> <!--"condition"-->
%<e>       <p><l>жай<s n="n"/></l>                 <r>яшен<s n="n"/></r></p></e> <!--"lightning"-->

\begin{figure*}[htbp]
\begin{texttt}
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{корал}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{морон}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{алы{\qipa ҫ}лы{\qipa ҡ}}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{йыра{\qipa ҡ}лы{\qipa ҡ}}<s n="n"/></r></p></e>
\end{texttt}
\caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}
\label{fig:bidix}
%\vspace{-1em}
\end{figure*}

% FIXME: the following sh... stuff needs to be rephrased
The bilingual lexicon currently contains 9,269 stem to stem correspondences and was build mostly by hand. Part of toponyms and other proper names was translated semi-automatically looking up in Wikipedia links (also some of the Russian loanwords were the result of intersection of Russian and Kazakh wordlists - "автомобиль" etc).
